A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-06-06 textos/optpaxos/draftv3.tex
draft v3: FTO-MCast

\newcommand{\blankline}{\vspace{4 mm}}
\newcommand{\cms}{\mbox{GF-Mcast}}
\newcommand{\cmsend}[1]{\mbox{\cms({#1})}}
\newcommand{\tconsm}{T_{cons}}
\newcommand{\tcons}{\mbox{$\tconsm$}}
\newcommand{\opt}{\mbox{GF-OPT-Deliver}}
\newcommand{\cons}{\mbox{GF-CONS-Deliver}}
\newcommand{\rmc}{\mbox{R-MCast}}
\newcommand{\rmd}{\mbox{R-Deliver}}
\newcommand{\optdel}[1]{\mbox{\opt({#1})}}
\newcommand{\consdel}[1]{\mbox{\cons({#1})}}
\newcommand{\rmcast}[2]{\mbox{\rmc({#1},{#2})}}
\newcommand{\rmdel}[1]{\mbox{\rmd({#1})}}

\title{Practical consistency management for geographically distributed MMOG servers}

@@

@@

We assume a system composed of nodes, divided into \emph{players} and \emph{servers}, distributed over a geographical area. Nodes may fail by crashing and subsequently recover, but do not experience arbitrary behavior (i.e., no Byzantine failures). Communication is done by message passing, through the primitives \emph{send}$(p,m)$ and \emph{receive}$(m)$, where $p$ is the addressee of message $m$. Messages can be lost but not corrupted. If a message is repeatedly resubmitted to a \emph{correct node} (defined below), it is eventually received.

Our protocols ensure safety under both asynchronous and synchronous execution periods. The FLP impossibility result~\cite{fischer1985idc} states that under asynchronous assumptions consensus cannot be both safe and live. We thus assume that the system is initially asynchronous and eventually becomes synchronous. The time when the system becomes synchronous is called the \emph{Global Stabilization Time (GST)}~\cite{dwork1988cpp}, and it is unknown to the nodes.
Before GST, there are no bounds on the time it takes for messages to be transmitted and actions to be executed. After GST, such bounds exist but are unknown. After GST nodes are either \emph{correct} or \emph{faulty}. A correct node is operational ``forever" and can reliably exchange messages with other correct nodes. This assumption is only needed to prove liveness properties about the system. In practice, ``forever" means long enough for one instance of consensus to terminate.

A game is composed of a set of objects. The game state is defined by the individual states of each one of its objects. We assume that the game objects are partitioned among different servers. Since objects have a location in the game, one way to perform this partitioning is by zoning the virtual world of the game. Each partition consists of a set of objects of the game and the server responsible for them is their \emph{coordinator}. As partitions represent physical regions in the game world, we define \emph{neighbor} regions as those which share a border with each other. We consider that each server is replicated, thus forming several groups which consist of the coordinator and its replicas. Therefore, for each region of the game world, there is a group assigned to it. This way, a group is said to coordinate an object when the group's coordinator is the coordinator for that object. Finally, groups are called neighbors when they are assigned to neighbor regions. From now on, the word `server' will be used for any kind of server, be it a coordinator or a replica.

Each player may send his command to one of the servers -- which might not be the group's coordinator, if that provides a lower delay between the issuing of a command and its delivery. In the case of avatar based virtual environments, and as an avatar is usually also an object, the server to which a player is connected belongs to the group of his avatar's coordinator. A command $C = \{ c_1, c_2, ... \}$ is composed of one or more subcommands, one subcommand per object it affects. We refer to the set of objects affected by command $C$ as $obj(C)$. Also, we refer to the objects coordinated by a server $S$ as $obj(S)$. Finally, we define $obj(C,S)$ = \{$o$ : $o \in obj(C)$ and $o \in obj(S)$\}.

@@

Our consistency criterion is ``eventual linearizability". (I'm not sure this is indeed what we want and how to define it, but we do need some consistency criterion...)

@@

\section{Protocol}

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with consensus (e.g. Paxos \cite{lamport1998ptp}). Each player sends his commands to the server to which he is connected, which then proposes that command in a consensus instance. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values.

However, providing such level of consistency may prove to be costly in an MMOG context, since there may be several communication steps between the sending of a command and its atomic delivery. For this reason, we use a primitive we call quasi-genuine global fifo total order multicast, which is described in section \ref{sec:quasi}.

\subsection{Quasi-genuine Global FIFO Multicast}
\label{sec:quasi}

To reduce the time needed to deliver a message, we use a multicast primitive which delivers messages optmistically, based on the time when they are created in a sender. This \emph{optimistic} delivery, although not guaranteeing that all replicas will receive all messages, is designed to have a fairly low latency, counting from when each message is sent until it is delivered.

The final -- conservative, fault tolerant, but costly in terms of communication steps for each message -- delivery order should be as close as possible to the optimistic one, so that no rollbacks would be deemed necessary. To explain how this works, we first should understand the idea behind the optimistic delivery. Also, we must define what ``quasi-genuine'' and ``global FIFO'' means.

@@

As messages are delivered according to their generation time, the delivery order is FIFO. Besides, as every two messages which are delivered in different groups should be delivered in the same order in these groups, we need total order. However, even FIFO total order is not enough to guarantee one property that we need: let the send time $st(m)$ of a message $m$ be the value of the wallclock of $p$ when $m$ was sent, where $p$ is the process which sent it. We need that, if $st(m_1) < st(m_2)$, then $m_1$ should be delivered before $m_2$, if $m_1$ and $m_2$ are delivered by the same process. Although this property may be hard to observe, it is used by the optimistic delivery protocol, so that messages can be delivered in order in one communication step, in the best case. If, when each process sends a message $m$, it stamps the message with the current value $st(m)$ of its wallclock, we define:

\textbf{Global FIFO order}: if processes $p_i$ and $p_j$ send respectively $m_i$ and $m_j$ to the same destination $p_d$, then if $st(m_i) < st(m_j)$, $m_i$ is not delivered after $m_j$ in $p_d$.

This property implies the FIFO order, although it does not guarantee that, if a message $m$ was sent, it will ever be delivered. As for the other properties of the algorithm, assuming a crash-stop model, we have:

@@

Regarding the \textbf{termination} property, as presented in \cite{rodrigues2000atomic}, for being guaranteed by the \cms\ primitive, it requires that an optimistic assumption is true:

\begin{center}
\emph{A2: every process $p$ knows a value $w(p)$, which is at least the maximum sum of the message delay bound plus the clock deviation between $p$ and any possible sender process $p'$ which could send a message to $p$.}
\end{center}

We then relaxed the termination property to another one, which considers the optimistic assumption we just stated:

\textbf{Optimistic termination}: For any message $m$, if a correct process \cms{}s $m$ \textbf{and the assumption A2 is true} then all correct processes that are destinations of $m$ \cons\ $m$.

@@

\subsubsection{Optimistic delivery}

@@

The basic idea of the optimistic ordering is the following: assuming that the processes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enough, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a process $p$ receives a message $m$ from a client, it immediately applies a timestamp $ts$ to it, which consists simply of the current value of $p$'s wallclock, $now$. Therefore, $m.ts=now$. A wait window of length $w(p)$ is considered, where $w(p)$ is defined as the highest value of the estimated communication delay plus the wallclock deviation between the process $p$ and any of the other processes in its group $G$ or in any of the groups in $sendersTo(G)$.

More formally, let $\delta(p',p)$ be the maximum time for a message from $p'$ to arrive at $p$. Also, let $\epsilon(p',p)$ be the deviation between the wallclocks of $p$ and $p'$. The value of $w(p)$ is estimated as the maximum value of \mbox{$(\delta(p',p) + \epsilon(p',p))$} for every $p'$ in $G = group(p)$ or in some group of $sendersTo(G)$.

After applying the timestamp to $m$, the process $p$ immediately forwards it to all the other processes involved, including those in other groups. A process is involved with a message $m$ when it is one of its destinations, which can be inferred from $m.dst$. Then, $p$ puts $m$ in an \mbox{\textit{optPending}} list, where it stays until $now>m.ts+w(p)$, which means that $m$ has been in that list for a time longer than the defined wait window. In the meantime, other messages, sent from other processes, may have been received and also inserted in that list, always sorting by their timestamps.

If $w(p)$ has been correctly estimated, and no message was lost, than all the messages that were supposed to be delivered before $m$ have necessarily been received already. If the same has occurred for all the processes, then all of them have received all the messages, and can deliver them in the same order of $ts$.

To avoid out-of-order deliveries, if when $m$ arrives at $p$, \mbox{$m.ts < now - w(p)$}, wich means that $m$ arrived too late, the message $m$ is simply discarded by $p$, since another message $m' : m'.ts > m.ts$ may have already been delivered\footnote{We could make this in a way such that $m$ is only discarded by $p$ if, in fact, there was a delivered message $m' : m'.ts > m.ts$. If there was no such message, $m$ could still be delivered without violating the order we defined.}.

However, even if the optimistic order is the same for all the processes, it won't be valid if the conservative order is different from it. For that reason, we devised a way to make the conservative delivery order as close as possible to the optimistic one.

\subsubsection{Conservative delivery}

Instead of having the conservative delivery done completely independently from the optimistic one, we can actually use the latter as a hint for the final delivery order. Since the optimistic delivery should be fast when compared to the conservative one, waiting for it should not decrease the system performance significantly. Also, if we wait a short period longer, we can avoid a rollback later caused by mismatches between the two delivery orders, which is not desirable. The basic idea is, then, to pick the optimistic delivery order seen at the processes of each group.

@@

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption A1, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that it would send no more messages with a timestamp lower than $t$ to group $G_{recv}$. We have defined that $sendersTo(G) =$ \mbox{$\{G' : G'\text{ is able to send a message to }G\}$}. When a process $p$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $p$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering (the ordering of messages originated in $G$) is done, all the messages with timestamp up to $t$ can be conservatively delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a process has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

We use consensus to conservatively deliver messages. Consider that each consensus instance $I$ from each group $I.grp$ receives a monotonically increasing unique integer identifier, without gaps, that is, for any two instances $I_i$ and $I_k$, such that $I_i.grp = I_k.grp$, if $I_i.id + 1 < I_k.id$, there is necessarily an instance $I_j : I_i.grp = I_j.grp = I_k.grp \wedge I_i.id < I_j.id < I_k.id$. No group runs two consensus instances in parallel: before initiating an instance of id $k+1$ each process checks whether the instance $k$ has already been decided, so some messages may wait to be proposed. When a process is allowed to initiate a new consensus instance, the pending messages may be proposed as a batch.%Consider also that each message $m$ sent by a group $G$ has a group sequence number $m.gs$ related to the order in which it is conservatively delivered, relatively to other messages also sent by $G$. As messages are conservatively delivered via consensus, the group sequence number of a message is equal to the id of the instance in which it was decided, that is, $m.gs = i \Leftrightarrow \exists I : I.id = i \wedge I.grp = m.src \wedge I.val = m$, where $I.val$ is the value decided by the instance $I$.

@@

  In this case, when $m$ is optimistically delivered by $p$, $p$ checks whether the latest consensus instance $I_{prev}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it, so other tasks can keep being executed. Then, once $I_{prev}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have either been already decided, or that have a timestamp smaller than that of some already decided message, are discarded from $propPending$. The rest is proposed as a batch in the consensus instance. Once $m$ is decided, it is not immediately delivered to the application. Instead, it is inserted in a $consPending$ list for later being conservatively delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

  In this case, when $now > m.ts + w(p)$, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $propPending$ that have either been already decided, or that have a timestamp smaller than that of some already decided message, are discarded from $propPending$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, if $G \in m.dst$, $m$ is inserted in the $consPending$ list. Besides, when $m$ is decided, $p$ sends a message $\{m, \text{`cons'}\}$ to every \mbox{$p' : (\exists G' \in m.dst \setminus \{G\}) \wedge (p' \in G')$}. When $\{m, \text{`cons'}\}$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $consPending$ list. If not, $p'$ inserts $m$ into $consPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $\{m, \text{`cons'}\}$ is received, every message \mbox{$\{m', \text{`cons'}\}:m'.ts<m.ts$} has already been received from $m.src$, every $\{m, \text{`cons'}\}$ message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel.}.

@@

  In this case, when $m$ is optimistically delivered by a process $p$ of group $G$, nothing else is done. The message $m$ is inserted into the \mbox{\textit{consPending}} list of $p$ only when $p$ receives some message $\{m, \text{`cons'}\}$.
\end{itemize}

The messages in the \textit{consPending} list are always sorted in ascending order of their timestamps. When the first message $m$, in the $consPending$ list of a process $p : group(p) = G$, is such that $m.ts < barrier(G, G')$ for all $G' \in sendersTo(G)$, then $m$ is conservatively delivered by $p$ to the application as the next message. We claim that this delivery respects the ``global FIFO total order''\footnote{Proof needed.}.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:quasi}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(k, val)$, which proposes a value $val$ for the consensus instance of id $k$ within its group; and also $Decide(k, val)$, which is called when the consensus instance of id $k$ finishes. $Decide(k, val)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $Decide(k, val)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a reliable multicast primitive \rmcast{$m$}{$groupSet$}, which \rmd{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step (e.g. the one described in \cite{urmcast1delta}).

Moreover, each process $p$ of group $G$ keeps some lists of messages:
\begin{itemize}
  \item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{propPending}, containing the messages waiting to be proposed by $p$;
  \item \textit{consPending}, with the messages ready to be \textit{\cons{}ed} (conservatively delivered), but which may be waiting for barriers from the groups in $sendersTo(G)$;
  \item \textit{decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{delivered}, which contains the messages that have been \cons{}ed already.
\end{itemize}

Also, each message $m$ has at least three fields: $dst$, which is the set of $m$'s destination groups; $src$, which is the group where $m$ was generated; and $ts$, which is the value of the wallclock of the process when $m$ was generated. %; and $gs$, which is $m$'s sequence number relatively to its group of origin. 
Note that, by abuse of notation, we have `process \mbox{$p \in m.dst$'} instead of `\mbox{$\exists$ group $G \in m.dst : \text{process }p \in G$'}.

Something that must be noticed is that some messages might not have their source groups as a destination. Anyway, each message $m$ of this type still has to be agreed upon in it group of origin $G$, so that its order among other messages from $G$ may be decided and %$m.gs$ can be decided, and also for 
$m$ will be retrievable, even in the presence of failures.% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

@@

\blankline
\INDENT{\textit{When} inserting into $optPending$ a message $m : (m.src = G \wedge \exists G' \in m.dst : G' \neq G)$}
  \INDENT{\textbf{for all} $G' \in m.dst : G' \neq G$ \textbf{do}}
    \STATE $lastBarrierCreated(G') \leftarrow m.ts$
  \ENDINDENT
\ENDINDENT

>>>>

\newcommand{\blankline}{\vspace{4 mm}}
\newcommand{\cms}{\mbox{QGFTO-Mcast}}
\newcommand{\cmsend}[1]{\mbox{\cms({#1})}}
\newcommand{\tconsm}{T_{cons}}
\newcommand{\tcons}{\mbox{$\tconsm$}}
\newcommand{\opt}{\mbox{GF-OPT-Deliver}}
\newcommand{\cons}{\mbox{QGFTO-Deliver}}
\newcommand{\rmc}{\mbox{FR-MCast}}
\newcommand{\rmd}{\mbox{FR-Deliver}}
\newcommand{\optdel}[1]{\mbox{\opt({#1})}}
\newcommand{\consdel}[1]{\mbox{\cons({#1})}}
\newcommand{\rmcast}[2]{\mbox{\rmc({#1},{#2})}}
\newcommand{\rmdel}[1]{\mbox{\rmd({#1})}}

\title{A Quasi-Genuine FIFO Total Order Multicast Primitive}

@@

Some multicast primitives have been devised in a such a way that multicast groups needed to communicate only when they had messages to exchange. These multicast primitives are called \emph{genuine}. We argue that it is possible, however, to devise a multicast primitive that, although not genuine, can make use of some knowledge given by the application to figure out which groups \emph{can} communicate with a given group. With such knowledge, although message exchanges take place even when there is no application message being transmitted between some two groups, such exchanges happen only when they are able to send to -- or receive from -- one another. The primitive that makes use of such property we call \emph{quasi-genuine}.

@@

We assume a system composed of nodes distributed over a geographical area. Nodes may fail by crashing and subsequently recover, but do not experience arbitrary behavior (i.e., no Byzantine failures). Communication is done by message passing, through the primitives \emph{send}$(p,m)$ and \emph{receive}$(m)$, where $p$ is the addressee of message $m$. Messages can be lost but not corrupted. If a message is repeatedly resubmitted to a \emph{correct node} (defined below), it is eventually received.

Our protocols ensure safety under both asynchronous and synchronous execution periods. The FLP impossibility result~\cite{fischer1985idc} states that under asynchronous assumptions consensus cannot be both safe and live. We thus assume that the system is initially asynchronous and eventually becomes synchronous. The time when the system becomes synchronous is called the \emph{Global Stabilization Time (GST)}~\cite{dwork1988cpp}, and it is unknown to the nodes.

Before GST, there are no bounds on the time it takes for messages to be transmitted and actions to be executed. After GST, such bounds exist but are unknown. After GST nodes are either \emph{correct} or \emph{faulty}. A correct node is operational ``forever" and can reliably exchange messages with other correct nodes. This assumption is only needed to prove liveness properties about the system. In practice, ``forever" means long enough for one instance of consensus to terminate.

%A game is composed of a set of objects. The game state is defined by the individual states of each one of its objects. We assume that the game objects are partitioned among different servers. Since objects have a location in the game, one way to perform this partitioning is by zoning the virtual world of the game. Each partition consists of a set of objects of the game and the server responsible for them is their \emph{coordinator}. As partitions represent physical regions in the game world, we define \emph{neighbor} regions as those which share a border with each other. We consider that each server is replicated, thus forming several groups which consist of the coordinator and its replicas. Therefore, for each region of the game world, there is a group assigned to it. This way, a group is said to coordinate an object when the group's coordinator is the coordinator for that object. Finally, groups are called neighbors when they are assigned to neighbor regions. From now on, the word `server' will be used for any kind of server, be it a coordinator or a replica.

%Each player may send his command to one of the servers -- which might not be the group's coordinator, if that provides a lower delay between the issuing of a command and its delivery. In the case of avatar based virtual environments, and as an avatar is usually also an object, the server to which a player is connected belongs to the group of his avatar's coordinator. A command $C = \{ c_1, c_2, ... \}$ is composed of one or more subcommands, one subcommand per object it affects. We refer to the set of objects affected by command $C$ as $obj(C)$. Also, we refer to the objects coordinated by a server $S$ as $obj(S)$. Finally, we define $obj(C,S)$ = \{$o$ : $o \in obj(C)$ and $o \in obj(S)$\}.

@@

%Our consistency criterion is ``eventual linearizability". (I'm not sure this is indeed what we want and how to define it, but we do need some consistency criterion...)

@@

%\section{Protocol}

%To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with consensus (e.g. Paxos \cite{lamport1998ptp}). Each player sends his commands to the server to which he is connected, which then proposes that command in a consensus instance. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values.

%However, providing such level of consistency may prove to be costly in an MMOG context, since there may be several communication steps between the sending of a command and its atomic delivery. For this reason, we use a primitive we call quasi-genuine global fifo total order multicast, which is described in section \ref{sec:quasi}.

\section{Quasi-genuine Global FIFO Multicast}
\label{sec:quasi}

%To reduce the time needed to deliver a message, we use a multicast primitive which delivers messages optmistically, based on the time when they are created in a sender. This \emph{optimistic} delivery, although not guaranteeing that all replicas will receive all messages, is designed to have a fairly low latency, counting from when each message is sent until it is delivered.

%The final -- conservative, fault tolerant, but costly in terms of communication steps for each message -- delivery order should be as close as possible to the optimistic one, so that no rollbacks would be deemed necessary. To explain how this works, we first should understand the idea behind the optimistic delivery. Also, we must define what ``quasi-genuine'' and ``global FIFO'' means.

@@

As messages are delivered according to their generation time, the delivery order is FIFO. As every two messages which are delivered in different groups should be delivered in the same order in these groups, we need total order. For the formal definition of the properties of the algorithm, assuming a crash-stop model, we have:

\textbf{Uniform Validity}: if a process \cms{}s $m$, then one of the correct processes that is a destination of $m$ eventually \cons{}s $m$.

@@

\textbf{FIFO Order}: if a correct process \cms{}s $m$ before it \cms{}s $m'$, then no correct process that is a destination of both $m$ and $m'$ \cons{}s $m'$, unless it has previously \cons{}ed  $m$.

@@

%\subsubsection{Optimistic delivery}

@@

%The basic idea of the optimistic ordering is the following: assuming that the processes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enough, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a process $p$ receives a message $m$ from a client, it immediately applies a timestamp $ts$ to it, which consists simply of the current value of $p$'s wallclock, $now$. Therefore, $m.ts=now$. A wait window of length $w(p)$ is considered, where $w(p)$ is defined as the highest value of the estimated communication delay plus the wallclock deviation between the process $p$ and any of the other processes in its group $G$ or in any of the groups in $sendersTo(G)$.

%More formally, let $\delta(p',p)$ be the maximum time for a message from $p'$ to arrive at $p$. Also, let $\epsilon(p',p)$ be the deviation between the wallclocks of $p$ and $p'$. The value of $w(p)$ is estimated as the maximum value of \mbox{$(\delta(p',p) + \epsilon(p',p))$} for every $p'$ in $G = group(p)$ or in some group of $sendersTo(G)$.

%After applying the timestamp to $m$, the process $p$ immediately forwards it to all the other processes involved, including those in other groups. A process is involved with a message $m$ when it is one of its destinations, which can be inferred from $m.dst$. Then, $p$ puts $m$ in an \mbox{\textit{optPending}} list, where it stays until $now>m.ts+w(p)$, which means that $m$ has been in that list for a time longer than the defined wait window. In the meantime, other messages, sent from other processes, may have been received and also inserted in that list, always sorting by their timestamps.

%If $w(p)$ has been correctly estimated, and no message was lost, than all the messages that were supposed to be delivered before $m$ have necessarily been received already. If the same has occurred for all the processes, then all of them have received all the messages, and can deliver them in the same order of $ts$.

%To avoid out-of-order deliveries, if when $m$ arrives at $p$, \mbox{$m.ts < now - w(p)$}, wich means that $m$ arrived too late, the message $m$ is simply discarded by $p$, since another message $m' : m'.ts > m.ts$ may have already been delivered\footnote{We could make this in a way such that $m$ is only discarded by $p$ if, in fact, there was a delivered message $m' : m'.ts > m.ts$. If there was no such message, $m$ could still be delivered without violating the order we defined.}.

%However, even if the optimistic order is the same for all the processes, it won't be valid if the conservative order is different from it. For that reason, we devised a way to make the conservative delivery order as close as possible to the optimistic one.

\subsection{Message delivery}

%Instead of having the conservative delivery done completely independently from the optimistic one, we can actually use the latter as a hint for the final delivery order. Since the optimistic delivery should be fast when compared to the conservative one, waiting for it should not decrease the system performance significantly. Also, if we wait a short period longer, we can avoid a rollback later caused by mismatches between the two delivery orders, which is not desirable. The basic idea is, then, to pick the optimistic delivery order seen at the processes of each group.

@@

Each message $m$ has a source group $m.src$, a set of destination groups $m.dst$ and a timestamp $m.ts$. Note that, by abuse of notation, we have `process \mbox{$p \in m.dst$'} instead of `\mbox{$\exists$ group $G \in m.dst : \text{process }p \in G$'}. The total order delivery of messages in a group can be solved by using consensus. Each consensus instance agrees upon some message set as the next ones to be delivered -- messages within the same set are told apart by the timestamp $m.ts$ applied by the process which created them; to solve timestamp collisions, the unique id of the sender process can be used. If processes send messages to each other using FIFO reliable channels, and after receival, messages are proposed via consensus in the order in which they are received, the FIFO delivery order is ensured.

The complicating factor is the possibility of a message having at least one destination group different from its source group. So, all involved groups must somehow agree regarding the delivery order of these messages. However, from assumption A1, each group knows which other groups it could send messages to -- or receive messages from. We can use this by defining \emph{barriers} for multicast, such that $barrier(G_{send},G_{recv})$ = $t$ means that the group $G_{send}$ promised that it would send no more messages with a timestamp lower than $t$ to group $G_{recv}$. We have defined that $sendersTo(G) =$ \mbox{$\{G' : G'\text{ is able to send a message to }G\}$}. When a process $p$, from group $G$, has received all the barrier values from all the groups in $sendersTo(G)$, and they are all greater than a value $t$, then $p$ knows that no more messages with timestamp lower than $t$ are coming from other groups and that, once the local ordering (the ordering of messages originated in $G$) is done, all the messages with timestamp up to $t$ can be delivered. Besides, a barrier is sent along with the bundle of all messages with timestamp greater than the last previous barrier sent from $G_{send}$ to $G_{recv}$, so that when a process has received a barrier from a group, it means that it knows all the messages sent by that group until the time value stored in that barrier.

As mentioned before, we use consensus to deliver messages. Consider that each consensus instance $I$ from each group $I.grp$ receives a monotonically increasing unique integer identifier, without gaps, that is, for any two instances $I_i$ and $I_k$, such that $I_i.grp = I_k.grp$, if $I_i.id + 1 < I_k.id$, there is necessarily an instance $I_j : I_i.grp = I_j.grp = I_k.grp \wedge I_i.id < I_j.id < I_k.id$. No group runs two consensus instances in parallel: before initiating an instance of id $k+1$ each process checks whether the instance $k$ has already been decided, so some messages may wait to be proposed. When a process is allowed to initiate a new consensus instance, the pending messages may be proposed as a batch.%Consider also that each message $m$ sent by a group $G$ has a group sequence number $m.gs$ related to the order in which it is conservatively delivered, relatively to other messages also sent by $G$. As messages are conservatively delivered via consensus, the group sequence number of a message is equal to the id of the instance in which it was decided, that is, $m.gs = i \Leftrightarrow \exists I : I.id = i \wedge I.grp = m.src \wedge I.val = m$, where $I.val$ is the value decided by the instance $I$.

@@

  In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{prev}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it, so other tasks can keep being executed. Then, once $I_{prev}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in the consensus instance. Once $m$ is decided, it is not immediately delivered to the application. Instead, it is inserted in a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

  In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, if $G \in m.dst$, $m$ is inserted in the $barPending$ list. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' : (\exists G' \in m.dst \setminus \{G\}) \wedge (p' \in G')$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel.}.

@@

  In this case, when $m$ is received for the first time\footnote{Multiple processes may have sent $m$. To ensure integrity and order, only the first delivery is considered.} by $p$, $m$ is inserted into the $barPending$ list of $p$ and the value of $barrier(G',G)$ is set to $m.ts$.
\end{itemize}

The messages in the \textit{barPending} list are always sorted in ascending order of their timestamps. When the first message $m$, in the $barPending$ list of a process $p : group(p) = G$, is such that $m.ts < barrier(G, G')$ for all $G' \in sendersTo(G)$, then $m$ is conservatively delivered by $p$ to the application as the next message. We claim that this delivery respects the FIFO total order\footnote{Proof needed.}.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:quasi}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(k, val)$, which proposes a value $val$ for the consensus instance of id $k$ within its group; and also $Decide(k, val)$, which is called when the consensus instance of id $k$ finishes. $Decide(k, val)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $Decide(k, val)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a FIFO reliable multicast primitive \rmcast{$m$}{$groupSet$}, which \rmd{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in FIFO order (e.g. the one described in \cite{ufrmcast1delta}).

Moreover, each process $p$ of group $G$ keeps some lists of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{propPending}, containing the messages waiting to be proposed by $p$;
  \item \textit{barPending}, with the messages ready to be \textit{\cons{}ed}, but which may be waiting for barriers from the groups in $sendersTo(G)$;
  \item \textit{decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{delivered}, which contains the messages that have been \cons{}ed already.
\end{itemize}

%; and $gs$, which is $m$'s sequence number relatively to its group of origin.

Something that must be noticed is that some messages might not have their source groups as a destination. Anyway, each message $m$ of this type still has to be agreed upon in its group of origin $G$, so that its order among other messages from $G$ may be decided and %$m.gs$ can be decided, and also for 
$m$ will be retrievable, even in the presence of failures.% Because of that, to ensure that messages of this kind are also received by $group(p)$'s coordinator $c$, which then initiates the Paxos instance that allows for the conservative delivery of the message in all of its destinations, every message $m$ from $p$ must also be sent to $c$ (l. \ref{algline:sendcoord}). This way, when the clock value at the coordinator $time(c) > w(c) + m.ts$, $c$ will conservatively send $m$ to all the processes in $m.dst$. Finally, to avoid delivering a message $m$ in a group it is not addressed to, before OPT-Delivering it, each process must check whether it is one of $m$'s destinations or not (l. \ref{algline:notinterested}). In l. \ref{algline:sendcoord}, $m$ is sent to the whole group $G$, so that in the case of failure of $G$'s coordinator, it is more likely that the new coordinator will know about the existence of $m$.

@@

\blankline
\INDENT{\textit{When} \rmc{}ing a message $m$ to some group $G' \neq G$}
  \INDENT{\textbf{for all} $G' \in m.dst : G' \neq G$ \textbf{do}}
    \STATE $lastBarrierCreated(G') \leftarrow m.ts$
  \ENDINDENT
\ENDINDENT
